---
output: 
  github_document:
    toc: true
    toc_depth: 3
    df_print: "tibble"
    includes:
      in_header: header.md
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "50%",
  cache = FALSE,
  autodep = TRUE,
  dpi = 72
)
```
    
## Installation

tinyVAST can be installed from GitHub:

```{r, eval=FALSE}
library(devtools)
install_github("vast-lib/tinyVAST", dependencies = TRUE)
```


## Citation

To cite tinyVAST in publications use:

```r
citation("tinyVAST")
```

Thorson, J. T., Anderson, S. C., Goddard, P., & Rooper, C. N. (2024).
tinyVAST: R package with an expressive interface to specify lagged
and simultaneous effects in multivariate spatio-temporal models
(arXiv:2401.10193). arXiv. <http://arxiv.org/abs/2401.10193>



## Related software

tinyVAST is builds upon many packages. This includes [VAST](https://github.com/James-Thorson-NOAA/VAST) R package:

Thorson, J.T. 2019. Guidance for decisions using the Vector Autoregressive Spatio-Temporal (VAST) package in stock, ecosystem, habitat and climate assessments. Fisheries Research 210: 143–161. <https://doi.org/10.1016/j.fishres.2018.10.013>.

and [sdmTMB](https://github.com/pbs-assess/sdmTMB) R package:

Anderson, S.C., E.J. Ward, P.A. English, L.A.K. Barnett. 2022. sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields. bioRxiv 2022.03.24.485545; doi: https://doi.org/10.1101/2022.03.24.485545

and the [glmmTMB](https://github.com/glmmTMB/glmmTMB) R package:

Brooks, M.E., Kristensen, K., van Benthem, K.J., Magnusson, A., Berg, C.W., Nielsen, A., Skaug, H.J., Maechler, M., and Bolker, B.M. 2017. glmmTMB balances speed and flexibility among packages for zero-inflated generalized linear mixed modeling. The R Journal 9(2): 378–400. <https://doi.org/10.32614/rj-2017-066>.

[INLA](https://www.r-inla.org/) and [inlabru](https://sites.google.com/inlabru.org/inlabru) can fit many of the same models as sdmTMB (and many more) in an approximate Bayesian inference framework.

[mgcv](https://cran.r-project.org/package=mgcv) can fit similar SPDE-based Gaussian random field models with code included in [Miller et al. (2019)](https://doi.org/10.1007/s13253-019-00377-z).

