---
title: "Seasonal index standardization"
author: "James T. Thorson"
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Seasonal index standardization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{sf}
---

```{r, include = FALSE}
has_lattice = requireNamespace("lattice", quietly = TRUE)
has_pdp = requireNamespace("pdp", quietly = TRUE)
EVAL <- has_lattice && has_pdp
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = EVAL,
  purl = EVAL
)
start_time = Sys.time()
# Install locally
#  devtools::install_local( R'(C:\Users\James.Thorson\Desktop\Git\tinyVAST)', force=TRUE )
# Build
#  setwd(R'(C:\Users\James.Thorson\Desktop\Git\tinyVAST)'); devtools::build_rmd("vignettes/multiple_data.Rmd"); rmarkdown::render( "vignettes/multiple_data.Rmd", rmarkdown::pdf_document())
```

```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
library(tinyVAST)
library(sf)
library(fmesher)
library(ggplot2)
```

`tinyVAST` can estimate seasonal spatio-temporal models using spatially varying coefficients [@thorson_spatially_2023] to incorporate additional spatial variation by season and then including an autoregressive spatio-temporal process across season-within-year indices.  This was originally demonstrated using package VAST [@thorson_seasonal_2020], and we here show code analyzing spring and fall bottom-trawl surveys for yellowtail flounder in the Northwest Atlantic conducted by the NEFSC, as well as a early-spring bottom trawl survey by DFO Canada.

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Load data
data( atlantic_yellowtail ) 

# Plot data extent
ggplot( atlantic_yellowtail ) + 
  geom_point( aes(x=longitude, y = latitude, col = season) ) +
  facet_wrap( vars(year) )
```

We then define a new year_season variable

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Define levels
atlantic_yellowtail$season = 
  factor(atlantic_yellowtail$season, levels = c("DFO", "SPRING", "FALL"))
atlantic_yellowtail$year_factor = 
  factor(atlantic_yellowtail$year, levels = sort(unique(atlantic_yellowtail$year)))

# Make season-year
atlantic_yellowtail$season_year = interaction(
  atlantic_yellowtail$season,
  atlantic_yellowtail$year_factor
)

# convert to an integer
atlantic_yellowtail$season_year_integer = match(
  atlantic_yellowtail$season_year,
  levels(atlantic_yellowtail$season_year)
)

# log-scale area swept
atlantic_yellowtail$log_swept = log(atlantic_yellowtail$swept)

# Add variable column
atlantic_yellowtail$var = "density"
```

We then define argument `spatial_varying` as a formula containing the space-season effect, and in this instance it replaces the `space_term`:

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Define mesh
mesh = fm_mesh_2d( 
  atlantic_yellowtail[,c('longitude','latitude')], 
  cutoff = 0.2 
)

# define formula with a catchability covariate for gear
formula = weight ~ season_year + offset(log_swept)

# Define SVC by season
spatial_varying = ~ season

spacetime_term = "
  density -> density, 1, ar_season
  density -> density, 3, ar_year
  density <-> density, 0, sd_spacetime
"

# fit using tinyVAST
fit = tinyVAST( 
  data = droplevels(atlantic_yellowtail), 
  # Specification
  formula = formula,
  spatial_varying = spatial_varying, 
  spacetime_term = spacetime_term,
  spatial_domain = mesh,
  family = tweedie("log"),
  # Indexing
  space_columns = c("longitude",'latitude'),
  time_column = "season_year_integer",
  times = unique(atlantic_yellowtail$season_year_integer),
  variable_column = "var",
  # Settings
  control = tinyVASTcontrol(
    trace = 1,
    profile = "alpha_j"
  )
)
```

We can inspect output, and see that both season-within-year (lag-1) and season-across-year (lag-3) terms are significant:

```{r}
summary(fit, "spacetime_term")
```


```{r, include = FALSE, warning=FALSE, message=FALSE}
run_time = Sys.time() - start_time
```
Runtime for this vignette: `r paste( round(unclass(run_time),2), attr(run_time, "units") )`
