---
title: "tinyVAST model description"
author: "James T. Thorson"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{tinyVAST model description}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: 'tinyVAST vignettes.bib'
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
# Install locally
#  devtools::install_local( R'(C:\Users\James.Thorson\Desktop\Git\tinyVAST)', force=TRUE )
# Build and PDF
#  setwd(R'(C:\Users\James.Thorson\Desktop\Git\tinyVAST)'); devtools::build_rmd("vignettes/model-description.Rmd"); rmarkdown::render( "vignettes/model-description.Rmd", rmarkdown::pdf_document())
#
# Recommended workflow:
#  * Open in Rstudio and knit using button there
```

\newcommand{\B}{\mathbf}   <!-- math bold nonitalic -->  
\newcommand{\PL}{\mathrm}   <!-- math nonitalic -->   
\newcommand{\s}{\boldsymbol{s}}

# Bivariate generalized linear mixed model structure

tinyVAST [@thorson_tinyvast_2025] is a bivariate extension of a generalized linear mixed model (see Tables 1 and 2 for notation), which includes two linear predictors that each include four additive components:

1. *Spatial interactions among variables*:  The user can specify interactions among variables at a given site (or for spatially correlated latent variables) using arrow notation derived from path analysis, based on the interface from R package sem;

2. *Temporal interaction among variables*:  The user can specify simultaneous and lagged interactions among variables over time using an expanded arrow-and-lag notation that is derived from R package dsem, where these interactions the annual intercept for a given variable therefore apply uniformly for all locations.

3. *Spatio-temporal interactions among variables*:  The user can specify simultaneous and lagged interactions among variables over time, where these interactions occur on a site-by-site basis.

4. *Generalized additive model*:  The user specifies a formula for a generalized additive model (GAM) that is interpreted by R package mgcv.  If other model components are missing, tinyVAST estimates parameters that are similar to mgcv, with small differences resulting from different methods for parameter estimation;

These four components are assembled into two linear predictors:

$$
\begin{aligned}
p_{\PL 1,i} &= \underbrace{\B X_{\PL 1,i} \B \alpha_{\PL 1} + \B Z_{\PL 1,i} \B \gamma_{\PL 1}}_\text{formula} + \underbrace{\B A_{i} \B \Omega_{\PL 1,c[i]}}_\text{space_term} + \underbrace{\B D_{\PL 1,c[i],t[i]}}_\text{time_term} + \underbrace{\B A_i \B E_{\PL 1,c[i],t[i]}}_\text{spacetime_term} + \underbrace{\B W_{\PL 1,i} (\B A_{i} \B \Xi_{\PL 1})^{\PL T} }_\text{spatially_varying}  \\
p_{\PL 2,i} &= \underbrace{\B X_{\PL 2,i} \B \alpha_{\PL 2} + \B Z_{\PL 2,i} \B \gamma_{\PL 2}}_\text{formula} + \underbrace{\B A_{i} \B \Omega_{\PL 2,c[i]}}_\text{space_term} + \underbrace{\B D_{\PL 2,c[i],t[i]}}_\text{time_term} + \underbrace{\B A_i \B E_{\PL 2,c[i],t[i]}}_\text{spacetime_term} + \underbrace{\B W_{\PL 2,i} (\B A_{i} \B \Xi_{\PL 2})^{\PL T} }_\text{spatially_varying}
\end{aligned}
$$

where

-   $p_{\PL 1,i}$ is the first linear predictor;
-   $\B X_{\PL 1,i} \B \alpha_{\PL 1} + \B Z_{\PL 1,i} \B \gamma_{\PL 1}$ is the contribution of the GAM component specified using a formula parsed by package `mgcv`; 
-   $\B \alpha_{\PL 1}$ is the GAM fixed effects with associated design matrix $\B X_{\PL 1}$, where $\B X_{\PL 1,i}$ is the row of the design matrix for sample $i$;
-   $\B \gamma_{\PL 1}$ is the GAM random effects with associated design matrix $\B Z$, where $\B Z_{\PL 1,i}$ is the row of the design matrix;
-   $\B A_{i} \B \Omega_{\PL 1,c[i]}$ is contribution of the space-veriable interaction $\B \Omega$ projected to sample $i$, where $\B A$ is the interpolation matrix with dimension $I \times S$ such that row $\B A_{i}$ interpolates spatial vector $\B \Omega_{\PL 1,c[i]}$ for sample $i$ and variable $c[i]$;
-   $\B D_{\PL 1,c[i],t[i]}$ is the contribution of the time-variable interaction $\B D$ for sample $i$;
-   $\B A_i \B E_{\PL 1,c[i],t[i]}$ is the contribution of the space-variable-time interaction $\B E$ projected to sample $i$;
-   $\B W_{\PL 1,i} \B A_{i} \B \Xi_{\PL 1,c[i]}$ is the contribution of spatially varying coefficients, i.e., a zero-centered and spatially varying slope $\Xi_{\PL 1,c[i]}$ that is projected to the location of samples, where these local slopes are then multiplied by  user-supplied covariates in design matrix $W_{\PL 1,i}$ 

and  terms are defined similarly for the second linear predictor $p_{\PL 2,i}$ except using subscript-2. The linear predictors are then passed through a bivariate inverse-link function to specify the distribution for errors:

$$
y_i \sim f_{e[i]}( g_{e[i]}^{-1} (p_{\PL 1,i}, p_{\PL 2,i}), \theta_{e[i]} )
$$
 
where

-   $f_{e[i]}$ probability density or mass function for sample $i$;
-   $g_{e[i]}^{-1} (p_{\PL 1,i}, p_{\PL 2,i})$ is the bivariate inverse-link function that transforms linear predictors to the central tendancy parameters of the distribution;
-   $\theta_{e[i]}$ is the dispersion parameters for distribution $e[i]$ for sample $i$;

In the simple case, the distribution only requires a single linear predictor such that $\B p_{\PL 2} = \B 0$ by construction and drops out of the model.  In this case, the model collapses to a generalized linear mixed model.  For example we might have a log-link and Poisson distribution, such that it collapses to a log-linked Poisson GLMM, $y_i \sim \mathrm{Poisson}(e^{p_{\PL 1,i}})$.

However, tinyVAST can also handle a delta-model using either logit-log or Poisson-linked link functions:

$$
g_{e[i]}^{-1} (p_{\PL 1,i}, p_{\PL 2,i}) = ( \mu_{\PL 1,i}, \mu_{\PL 2,i}  )
$$

where $\mu_{\PL 1,i}$ and $\mu_{\PL 2,i}$ are the two linear predictors, such that $\mathbb{E}[y_i] = \mu_{\PL 1,i} \mu_{\PL 2,i}$;

For the conventional logit-log bivariate link function we use a logit-link for encounter probabilities and a log-link for positive catch rates:

$$
\begin{aligned}
\mu_{\PL 1,i} &= \frac{e^{p_{\PL 1,i}}}{1+e^{p_{\PL 1,i}}}  \\
\mu_{\PL 2,i} &= e^{p_{\PL 2,i}}
\end{aligned}
$$

while for the Poisson-linked link function [@thorson_three_2018] we specify a complemetary log-log link for encounter probabilities and define the second link function such that $\mu_{\PL 1,i} \mu_{\PL 2,i} = e^{p_{\PL 1,i}} e^{p_{\PL 2,i}}$:

$$
\begin{aligned}
\mu_{\PL 1,i} &= 1 - e^{-e^{p_{\PL 1,i}}} \\
\mu_{\PL 2,i} &= \frac{e^{p_{\PL 1,i}}}{\mu_{\PL 1,i}} e^{p_{\PL 2,i}}
\end{aligned}
$$
where $e^{p_{\PL 1,i}}$ is the density for an underlying point process, and $e^{p_{\PL 2,i}}$ is biomass per point (i.e., animal or group).  

In either the conventional or Poisson-linked delta model, $\mu_{\PL 1,i} = \mathrm{Pr}(Y>0)$ is the encounter probability (and is fitted with a Bernoulli distribution), while $\mu_{\PL 2,i}$ is the central tendancy for positive values, i.e., $\mu_{\PL 2,i} = \mathbb{E}(Y | Y>0)$.

Usefully, the interface allows analysts to specify a different distribution $e_i$ for different partitions of the data.  This allows the model to jointly analyze different data types [@gruss_developing_2019], or data with different magnitudes of sampling error.  

# Spatial domains

Linear predictors include spatially autocorrelated latent variables.  These variables are treated as Gaussian Markov random fields (GMRFs), and evaluating the probability density of GMRFs involves calculating the precision matrix $\B Q_{\PL{domain}} = \B \Sigma^{-1}$ as the inverse of the spatial covariance matrix [@thorson_spatio-temporal_2024].  tinyVAST involves three options for specifying this spatial precision:

### Stochastic partial differentiaul equation (SPDE)

The analyst can approximate spatial variation over a continuous two-dimensional surface by constructing finite element mesh (FEM), treating the value at vertices as a GMRF, and then using bilinear interpolation (i.e., a piecewise linear approximation) to interpolate from vertices to the spatial domain.  This method was developed by [@lindgren_explicit_2011], as popularized in the software R-INLA  [@lindgren_continuous_2012], first implemented in TMB by [@thorson_importance_2014], and using elements constructed by R-package `fmesher` [@lindgren_fmesher_2023]. In this case, the precision is constructed as:

$$
\B Q_{\PL{domain}} = \tau^2 ( \kappa^4 \B M_{\PL 0} + 2\kappa^2 \B M_{\PL 1} + \B M_{\PL 2} )
$$
where every row $\B A_i$ of the interpolation matrix $\B A$ is nonzero for only the three vertices of the triangle that contains sample $i$

### Simultaneous autoregressive (SAR):

Alternatively, the analyst can apply a simultaneous autoregressive process [@ver_hoef_relationship_2018], specifying an areal model that represents the value within spatial strata:

$$
\B Q_{\PL{domain}} = \tau^2 (\B I - \kappa \B A^*)^2 
$$
where $\B A^*$ is the adjacency matrix of the graph specified by the analyst, $\kappa$ is the estimated partial correlation between adjacent areas, and each row $\B A_i$ of interpolation matrix $\B A$ is nonzero only for the single spatial stratum containing sample $i$ (and noting that adjacency matrix $\B A^*$ is different from interpolation matrix $\B A$, but we use the same notation for both due to a collision in notational standards).

### Stream networks using a tail-down exponential correlation function

Finally, the analyst can specify that sites are partially correlated if they are adjacent along a stream network (while ignoring flow direction).  This results in an Onstein-Uhlenbeck process along the stream network [@charsley_catchment-scale_2023], or an *exponential tail-down* model:

$$
\B{Q}_{\PL{domain}} = \B{(I+D-P)}^T \B{I+D}^{-1} \B{(I+D-P)}
$$
where $\B D$ is a sparse diagonal matrix with diagonal elements 

$$
d_{i,j} =\PL{exp}(-2 \theta |\B s_i, \B s_j|) / (1 - \PL{exp}(-2 \theta |\B s_i, \B s_j|))
$$
where $|\B s_i, \B s_j|$ is the distance from downstream (parent) node $j$ to upstream (child) node $i$, and $\theta$ is the O-U parameter governing the decorrelation rate.  Similarly, $\B P$ is a sparse matrix containing values $\rho_{i,j}$, where:
$$
\rho_{i,j} = \PL{exp}(-\theta |\B s_i, \B s_j|) / (1 - \PL{exp}(-2 \theta |\B s_i, \B s_j|))
$$
such that $\rho_{i,j}$ the regression slope predicting upstream node $i$ from downstream node $j$.  The spatial interpolation matrix $\B A$ again has row $\B A_i$ for each sampling or predictive location, and $\B A_i$ is nonzero for the two nodes immediately up and downstream from a given location, with values such that a given location is predicted as a weighted average based upon the distance from those two nodes.  

For the SPDE and SAR spatial domains, the term $\tau$ is defined such that precision $\B Q_{\PL{domain}}$ has unit variance.  This is done because the spatial domain always arises in combination with other parameters, which are used to define the variance of the associated spatial variable.

# Structural equation models

tinyVAST also involves specifying a structural equation model (SEM).  This SEM be viewed either:

1. *Weak interpretation*: as an expressive interface to parameterize the correlation among variables, using as many or few parameters as might be appropriate; or
2. *Strong interpretation*: as a structural causal model, allowing predictions about the consequence of counterfactual changes to the system.

To specify a SEM, the user uses *arrow notation* derived from package `sem` [@fox_structural_2006], as described in TMB by [@thorson_spatio-temporal_2024].  For example, to specify a linear model this involves:

```{r echo=TRUE, eval=FALSE}
w1 -> w2, b_12
w1 <-> w1, sd_1
w2 <-> w2, sd_2
```

This then estimates a single slope parameter (represented with a one-headed arrow), as well as the variance of $W_1$ and $W_2$ (specified with two-headed arrows).  In a more complicated case, $W_1$ might cause $W_2$, which in turn causes $W_3$.  This is then represented as:

```{r arrow notation, echo=TRUE, eval=FALSE}
# Path, parameter_name, start value
w1 -> w2, b_12, 0
w2 -> w3, b_23, 0
w1 <-> w1, s_1, 1
w2 <-> w2, s_2, 1
w3 <-> w3, s_3, 1
```

SEM interactions can be as complicated or simple as desired, and can include:

1.  Latent variables and loops (i.e., they are not restricted to directed acyclic graphs);
2.  Values that are fixed a priori, where the `parameter_name` is provided as `NA` and the starting value that follows is the fixed value;
3.  Values that are mirrored among path coefficients, where the same `parameter_name` is provided for multiple rows of the text file.

In this preceding example, path coefficients for one-headed arrows then define path matrix $\B P$: 

$$
\B P = 
\begin{pmatrix}
  0 & 0 & 0 \\
  b_{12} & 0 & 0 \\
  0 & b_{23} & 0 
\end{pmatrix}
$$

and coefficents for two-headed arrows define the Cholesky $\B G$ of the exnogenous covariance matrix $\B G^T \B G$:

$$
\B G = 
\begin{pmatrix}
  s_{1} & 0 & 0 \\
  0 & s_{2} & 0 \\
  0 & 0 & s_{3} 
\end{pmatrix}
$$

These matrices are define a simultaneous equation model:

$$
\mathbf{ w = P w + \epsilon} \\
\B \epsilon \sim \mathrm{MVN}( \B 0, \B G^T \B G )
$$
where the variance $\mathrm{Var}(\B w) = (\mathbf{I - P})^{-1} \B G^2 (\mathbf{I - P}^T)^{-1}$.  This then results in a sparse precision matrix:

$$
\B Q = (\mathbf{I - P}^T) \B G^{-1} \B G^{-T} (\mathbf{I - P})
$$

where this precision matrix $\B Q$ is then used as a modular component in the larger tinyVAST model.  

# Dynamic structural equation models

Similarly, tinyVAST involves specifying dynamic structural equation models (DSEM) [@thorson_dynamic_2024].  To specify a DSEM, the user uses *arrow-and-lag notation*.  For example, to specify a univariate first-order autoregressive process:

```{r echo=TRUE, eval=FALSE}
w1 -> w1, 1, rho, 0.8
```

If there were four time-intervals ($T=4$) this would then result in the path matrix:

$$
\B P = 
\begin{pmatrix}
  0 & 0 & 0 & 0 \\
  \rho & 0 & 0 & 0 \\
  0 & \rho & 0 & 0 \\
  0 & 0 & \rho & 0 
\end{pmatrix}
$$

and when the DSEM involves multiple times and variables, the sparse precision is formed by summing across the Kronecker product of time-lag and interaction matrices.  This DSEM defines a GMRF over a nonseparable interaction of time and variables, represented by a matrix $\B Q_{\PL{time\_term}}$ with dimension $CT \times CT$.  The user can specify a separate *arrow-and-lag* notation to define the precision matrix for the time-variable interaction $\B Q_{\PL{time\_term}}$ and for the space-time-variable interaction $\B Q_{\PL{spacetime\_term}}$.  

The precision matrix $\B Q_{\PL{time\_term}}$ for the time term $\B D$ is used to define the time-varying intercept for each variable:
 
$$
\mathrm{vec}(\B D) \sim \mathrm{MVN}(\B 0, \B Q_{\PL{time\_term}})
$$
Meanwhile, the space-time term $\B Q_{\PL{spacetime\_term}}$ is combined with the spatial precision $\B Q_{\PL{space\_term}}$ as we explain next.

# Spatial interactions for SEM and DSEM

tinyVAST uses the SEM and DSEM notation to construct the joint precision for the space-variable interaction $\B \Omega$ with dimension $S \times C$, and the space-time-variable interaction $\B E$ with dimension $S \times C \times T$.  To do so, it constructs the separable precision for each process:

$$
\mathrm{vec}(\B E) \sim \mathrm{MVN}(\B 0, \B Q_{\PL{spacetime\_term}} \otimes \B Q_{\PL{domain}})
$$

where the precision matrix $\B Q_{\PL{spacetime\_term}} \otimes \B Q_{\PL{domain}}$ has dimension $STC \times STC$ to match the length of $\mathrm{vec}(\B E)$, and 
$$
\mathrm{vec}(\B \Omega) \sim \mathrm{MVN}(\B 0, \B Q_{\PL{space\_term}} \otimes \B Q_{\PL{domain}})
$$

where the precision matrix $\B Q_{\PL{space\_term}} \otimes \B Q_{\PL{domain}}$ has dimension $SC \times SC$, to match the length of $\mathrm{vec}(\B \Omega)$.  This specification generalizes spatial factor analysis [@thorson_spatial_2015] and spatial dynamic factor analysis [@thorson_joint_2016].

# Generalized additive model

The analyst can specify a generalized additive model using syntax from package `mgcv` [@wood_generalized_2006], and parsed into TMB using an interface developed by `sdmTMB` [@anderson_sdmtmb_nodate].  For example this might involve:

```{r echo=TRUE, eval=FALSE}
count ~ year + offset(log_area) + s(depth) + s(species, bs="re")
```

If `year` and `species` are factors and `depth` and `log_area` are continuous, then this would specify a fixed effect for each level `year`, a spline smoother for `depth`, using `log_area` as an offset, and estimating a random intercept for each level of `species`.  This formula is parsed internally to assemble fixed effects in design matrix $\B X$ and the basis functions for spline smoothers and random effects in design matrix $\B Z$.  The coefficients $\B \gamma$ associated with smoothers and random effects are then specified to follow a GMRF:

$$
\B \gamma \sim \mathrm{GMRF}( \B 0, \B Q_{\PL{gam}})
$$
where $\B Q_{\PL{gam}}$ is a blockwise diagonal matrix, assembled from estimated variance parameters and matrices constructed by mgcv.  

# Spatially varying coefficients

Finally, the analyst can account for spatial variation in the relationship between a specified covariate and the estimated response.  This is done by specifying a spatially varying coefficient (SVC) [@thorson_spatially_2023], and this can be used e.g. to account for spatially varying differences in catchability/detectability [@gruss_spatially_2023] or to account for nonlocal impacts of regional oceanographic indices [@thorson_measuring_2019].  This involves estimating each SVC $l \in \{1,2,...,L\}$ as a zero-centered Gaussian Markov random field while estimating its corresponding variance $\sigma_{\PL{\xi},l}^2$:

$$
\Xi_{\PL 1,l} \sim \mathrm{GMRF}( \B 0, \sigma_{\PL{\xi},l}^{-2} \B Q_{\PL{domain}} )
$$

Any covariate specified as an SVC it is also typically specified in the `formula` to estimate a non-zero mean for the SVC. If the estimated variance of the SVC approaches zero, it then suggests that the covariate slope does not vary spatially.  

# Notation summary

```{r indices}
subscripts <- tibble::tribble(
  ~Symbol, ~Description,
  "$i$", "Index for each sample, $i$ in $(1,2,...,I)$",
  "$s[i]$", "spatial coordinate for sample $i$, $s$ in $(1,2,...,S)$",
  "$t[i]$", "time-interval for sample $i$, $t$ in $(1,2,...,T)$",
  "$c[i]$", "category for sample $i$, $c$ in $(1,2,...,C)$",
  "$e[i]$", "error distribution and link function for sample $i$"
)
knitr::kable(subscripts, caption = "Table 1: Subscript notation", escape = FALSE, booktabs = TRUE)
```

```{r symbols, results='asis'}
symbols <- tibble::tribble(
   ~Symbol, ~Code, ~Description,
  "$y$", "`y_i`", "Observed response data",
  "$p_1$", "`p_i`", "first linear predictor",
  "$p_2$", "`p2_i`", "second linear predictor",
)
knitr::kable(symbols, caption = "Table 2: Symbol notation, code representation (in model output or in model template code), and descriptions.", escape = FALSE, booktabs = TRUE,
   linesep = c(
    rep('', 7), # y, mean etc.
    '\\addlinespace',
    rep('', 7), # fields
    '\\addlinespace',
    rep('', 5), # covariance stuff
    '\\addlinespace',
    rep('', 5), # more covariance stuff
    '\\addlinespace',
    rep('', 99) # end
  ))
```

# Works cited
