---
title: "Simultaneous autoregressive process"
author: "James T. Thorson"
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Simultaneous autoregressive process}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggraph2}
---

```{r, include = FALSE}
has_ggplot = requireNamespace("ggplot2", quietly = TRUE)
EVAL <- has_ggplot
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = EVAL,
  purl = EVAL
)
# Install locally
#  devtools::install_local( R'(C:\Users\James.Thorson\Desktop\Git\tinyVAST)', force=TRUE )
# Build
#  setwd(R'(C:\Users\James.Thorson\Desktop\Git\tinyVAST)'); devtools::build_rmd("vignettes/simultaneous_autoregressive_process.Rmd"); rmarkdown::render( "vignettes/simultaneous_autoregressive_process.Rmd", rmarkdown::pdf_document())
```

```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
library(tinyVAST)
library(igraph)
library(rnaturalearth)
library(sf)
options("tinyVAST.verbose" = FALSE)
```

`tinyVAST` is an R package for fitting vector autoregressive spatio-temporal (VAST) models using a minimal and user-friendly interface.
We here show how it can fit a multivariate second-order autoregressive (AR2) model including spatial correlations using
a simultaneous autoregressive (SAR) process specified using _igraph_.

To do so, we first load salmong returns, and remove 0s to allow comparison between Tweedie and lognormal distributions.

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
data( salmon_returns )

# Transform data
salmon_returns$Biomass_nozeros = ifelse( salmon_returns$Biomass==0,
                                         NA, salmon_returns$Biomass )
Data = na.omit(salmon_returns)
```

We first explore an AR2 process, with independent variation among regions.
This model shows a substantial first-order autocorrelation for sockeye and chum,
and substantial second-order autocorrelation for pink salmon.  An AR(2) process is
stationary if $\phi_1 + \phi_2 < 1$ and $\phi_2 - \phi_1 < 1$, and this
stationarity criterion suggests that each time-series is close to (but not quite) nonstationary.
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Define graph for SAR process
unconnected_graph = make_empty_graph( nlevels(Data$Region) )
V(unconnected_graph)$name = levels(Data$Region)
plot(unconnected_graph)

# Define SEM for AR2 process
dsem = "
  sockeye -> sockeye, -1, lag1_sockeye
  sockeye -> sockeye, -2, lag2_sockeye

  pink -> pink, -1, lag1_pink
  pink -> pink, -2, lag2_pink

  chum -> chum, -1, lag1_chum
  chum -> chum, -2, lag2_chum
"

# Fit tinyVAST model
mytiny0 = tinyVAST(
     formula = Biomass_nozeros ~ 0 + Species + Region,
     data = Data,
     dsem = dsem,
     variable_column = "Species",
     time_column = "Year",
     space_column = "Region",
     distribution_column = "Species",
     family = list( "chum" = lognormal(),
                          "pink" = lognormal(),
                          "sockeye" = lognormal() ),
     spatial_graph = unconnected_graph,
     control = tinyVASTcontrol( profile="alpha_j" ) )

# Summarize output
Summary = summary(mytiny0, what="dsem")
knitr::kable( Summary, digits=3)
```

We also explore an SAR process for adjacency among regions
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Define graph for SAR process
adjacency_graph = make_graph( ~ Korea - Japan - M.I - WKam - EKam -
                                WAK - SPen - Kod - CI - PWS -
                                SEAK - NBC - SBC - WA )
```

We can plot this adjacency on a map to emphasize that it is a simple way to encode information about spatial proximity:
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=7, fig.height=5}
#maps = ne_countries( country = c("united states of america","russia","canada","south korea","north korea","japan") )
maps = ne_countries( continent = c("north america","asia","europe") )
maps = st_combine( maps )
maps = st_transform( maps, crs=st_crs(3832) )
#maps = st_crop( maps, xmin = -5*1e5, xmax = 12*1e5,
#                      ymin = 0 * 1e6, ymax=10 * 1e6 )

# Format inputs
loc_xy = cbind(
  x = c(129,143,140,156,163,-163,-161,-154,-154,-147,-138,-129,-126,-125),
  y = c(36,40,57,53,57,60,55,56,59,61,57,54,50,45)
)
loc_xy = sf_project( loc_xy, from=st_crs(4326), to=st_crs(3832) )

# Plot
xlim = c(-4,10) * 1e6
ylim = c(3,10) * 1e6
plot( maps, 
      xlim = xlim,
      ylim = ylim,
      col = "grey",
      asp = FALSE,
      add = FALSE )
plot( adjacency_graph, 
      layout = loc_xy,
      add = TRUE, 
      rescale = FALSE,
      vertex.label.color = "red",
      xlim = xlim, 
      ylim = ylim, 
      edge.width = 2,
      edge.color = "red" )
```

We can then pass this adjacency graph to `tinyVAST` during fitting:
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Fit tinyVAST model
mytiny = tinyVAST(
     formula = Biomass_nozeros ~ 0 + Species + Region,
     data = Data,
     dsem = dsem,
     variable_column = "Species",
     time_column = "Year",
     space_column = "Region",
     distribution_column = "Species",
     family = list( "chum" = lognormal(),
                          "pink" = lognormal(),
                          "sockeye" = lognormal() ),
     spatial_graph = adjacency_graph,
     control = tinyVASTcontrol( profile="alpha_j" ) )

# Summarize output
Summary = summary(mytiny, what="dsem")
knitr::kable( Summary, digits=3)
```

We can use AIC to compare these two models.  This comparison suggests that spatial adjancency
is not a parsimonious way to describe correlations among time-series.
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# AIC for unconnected time-series
AIC(mytiny0)
# AIC for SAR spatial variation
AIC(mytiny)
```

Finally, we can plot observations and predictions for the selected model
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Compile long-form dataframe of observations and predictions
Resid = rbind( cbind(Data[,c('Species','Year','Region','Biomass_nozeros')], "Which"="Obs"),
               cbind(Data[,c('Species','Year','Region')], "Biomass_nozeros"=predict(mytiny0,Data), "Which"="Pred") )

# plot using ggplot
library(ggplot2)
ggplot( data=Resid, aes(x=Year, y=Biomass_nozeros, col=Which) ) + # , group=yhat.id
  geom_line() +
  facet_grid( rows=vars(Region), cols=vars(Species), scales="free" ) +
  scale_y_continuous(trans='log')  #
```
