---
title: "Spatial factor analysis"
author: "James T. Thorson"
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Spatial factor analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_fmesher = requireNamespace("fmesher", quietly = TRUE)
has_knitr = requireNamespace("knitr", quietly = TRUE)
EVAL <- has_fmesher && has_knitr
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = EVAL,
  purl = EVAL
)
# Install locally
#  devtools::install_local( R'(C:\Users\James.Thorson\Desktop\Git\tinyVAST)', force=TRUE )
# Build
#  setwd(R'(C:\Users\James.Thorson\Desktop\Git\tinyVAST)'); devtools::build_rmd("vignettes/spatial_factor_analysis.Rmd"); rmarkdown::render( "vignettes/spatial_factor_analysis.Rmd", rmarkdown::pdf_document())
```

```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
library(tinyVAST)
library(fmesher)
set.seed(101)
```

`tinyVAST` is an R package for fitting vector autoregressive spatio-temporal (VAST) models.
We here explore the capacity to specify a spatial factor analysis, where the spatial pattern for multiple variables is described via
their estimated association with a small number of spatial latent variables.

# Spatial factor analysis
We first explore the ability to specify two latent variables for five manifest variables.  To start we simulate two spatial latent variables,
project via a simulated loadings matrix, and then simulate a Tweedie response for each manifest variable:

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Simulate settings
theta_xy = 0.4
n_x = n_y = 10
n_c = 5
rho = 0.8
resid_sd = 0.5

# Simulate GMRFs
R_s = exp(-theta_xy * abs(outer(1:n_x, 1:n_y, FUN="-")) )
R_ss = kronecker(X=R_s, Y=R_s)
delta_fs = mvtnorm::rmvnorm(n_c, sigma=R_ss )

#
L_cf = matrix( rnorm(n_c^2), nrow=n_c )
L_cf[,3:5] = 0
L_cf = L_cf + resid_sd * diag(n_c)

#
d_cs = L_cf %*% delta_fs
```

Where we can inspect the simulated loadings matrix
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
dimnames(L_cf) = list( paste0("Var ", 1:nrow(L_cf)),
                       paste0("Factor ", 1:ncol(L_cf)) )
knitr::kable( L_cf,
              digits=2, caption="True loadings")
```

We then specify the model as expected by _tinyVAST_:

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Shape into longform data-frame and add error
Data = data.frame( expand.grid(species=1:n_c, x=1:n_x, y=1:n_y),
                   "var"="logn", "z"=exp(as.vector(d_cs)) )
Data$n = tweedie::rtweedie( n=nrow(Data), mu=Data$z, phi=0.5, power=1.5 )
mean(Data$n==0)

# make mesh
mesh = fm_mesh_2d( Data[,c('x','y')] )

#
sem = "
  f1 -> 1, l1
  f1 -> 2, l2
  f1 -> 3, l3
  f1 -> 4, l4
  f1 -> 5, l5
  f2 -> 2, l6
  f2 -> 3, l7
  f2 -> 4, l8
  f2 -> 5, l9
  f1 <-> f1, NA, 1
  f2 <-> f2, NA, 1
  1 <-> 1, NA, 0
  2 <-> 2, NA, 0
  3 <-> 3, NA, 0
  4 <-> 4, NA, 0
  5 <-> 5, NA, 0
"

# fit model
out = fit( sem = sem,
           data = Data,
           formula = n ~ 0 + factor(species),
           spatial_graph = mesh,
           family = tweedie(),
           variables = c( "f1", "f2", 1:n_c ),
           data_colnames = list( space = c("x","y"), variable = "species",
                                 time = "time", distribution = "dist"),
           control = tinyVASTcontrol(quiet=TRUE, trace=0, gmrf="proj") )
out
```

We can compare the true loadings (rotated to optimize comparison):

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
Lrot_cf = rotate_pca( L_cf )$L_tf
dimnames(Lrot_cf) = list( paste0("Var ", 1:nrow(Lrot_cf)),
                       paste0("Factor ", 1:ncol(Lrot_cf)) )
knitr::kable( Lrot_cf,
              digits=2, caption="Rotated true loadings")
```

with the estimated loadings

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Extract and rotate estimated loadings
Lhat_cf = matrix( 0, nrow=n_c, ncol=2 )
Lhat_cf[lower.tri(Lhat_cf,diag=TRUE)] = as.list(out$sdrep, what="Estimate")$theta_z
Lhat_cf = rotate_pca( L_tf=Lhat_cf, order="decreasing" )$L_tf
```

Where we can compared the estimated and true loadings matrices:
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
dimnames(Lhat_cf) = list( paste0("Var ", 1:nrow(Lhat_cf)),
                       paste0("Factor ", 1:ncol(Lhat_cf)) )
knitr::kable( Lhat_cf,
              digits=2, caption="Rotated estimated loadings" )
```


Or we can specify the model while ensuring that residual spatial variation is also captured:

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
#
sem = "
  f1 -> 1, l1
  f1 -> 2, l2
  f1 -> 3, l3
  f1 -> 4, l4
  f1 -> 5, l5
  f2 -> 2, l6
  f2 -> 3, l7
  f2 -> 4, l8
  f2 -> 5, l9
  f1 <-> f1, NA, 1
  f2 <-> f2, NA, 1
  1 <-> 1, sd_resid
  2 <-> 2, sd_resid
  3 <-> 3, sd_resid
  4 <-> 4, sd_resid
  5 <-> 5, sd_resid
"

# fit model
out = fit( sem = sem,
           data = Data,
           formula = n ~ 0 + factor(species),
           spatial_graph = mesh,
           family = list( "obs"=tweedie() ),
           variables = c( "f1", "f2", 1:n_c ),
           data_colnames = list( space = c("x","y"), variable = "species",
                                 time = "time", distribution = "dist"),
           control = tinyVASTcontrol(quiet=TRUE, trace=0, gmrf="proj") )

# Extract and rotate estimated loadings
Lhat_cf = matrix( 0, nrow=n_c, ncol=2 )
Lhat_cf[lower.tri(Lhat_cf,diag=TRUE)] = as.list(out$sdrep, what="Estimate")$theta_z
Lhat_cf = rotate_pca( L_tf=Lhat_cf, order="decreasing" )$L_tf
```

Where we can again compared the estimated and true loadings matrices:
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
dimnames(Lhat_cf) = list( paste0("Var ", 1:nrow(Lhat_cf)),
                       paste0("Factor ", 1:ncol(Lhat_cf)) )
knitr::kable( Lhat_cf,
              digits=2, caption="Rotated estimated loadings with full rank" )
```

