---
title: "Vector autoregressive spatio-temporal models"
author: "James T. Thorson"
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{VAST}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{pdp}
---

```{r, include = FALSE}
has_vast = requireNamespace("VAST", quietly = TRUE)
if( isTRUE(has_vast) ){
  updated_vast = isTRUE( packageVersion("VAST") >= package_version("3.11.0") )
}else{
  updated_vast = FALSE
}
has_sdmTMB = requireNamespace("sdmTMB", quietly = TRUE)
has_ggplot = requireNamespace("ggplot2", quietly = TRUE)
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && has_sdmTMB && has_vast && has_ggplot
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = EVAL,
  purl = EVAL
)
# Install locally
#  devtools::install_local( R'(C:\Users\James.Thorson\Desktop\Git\tinyVAST)', force=TRUE )
# Build
#  setwd(R'(C:\Users\James.Thorson\Desktop\Git\tinyVAST)'); devtools::build_rmd("vignettes/VAST.Rmd"); rmarkdown::render( "vignettes/VAST.Rmd", rmarkdown::pdf_document())
```

```{r setup, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
library(tinyVAST)
library(fmesher)
set.seed(101)
options("tinyVAST.verbose" = FALSE)
```

`tinyVAST` is an R package for fitting vector autoregressive spatio-temporal (VAST) models.
We here explore the capacity to specify the vector-autoregressive spatio-temporal component.

# Spatio-temporal autoregressive model
We first explore the ability to specify a first-order autoregressive spatio-temporal process:

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Simulate settings
theta_xy = 0.4
n_x = n_y = 10
n_t = 15
rho = 0.8
spatial_sd = 0.5

# Simulate GMRFs
R_s = exp(-theta_xy * abs(outer(1:n_x, 1:n_y, FUN="-")) )
V_ss = spatial_sd^2*kronecker(R_s, R_s)
d = mvtnorm::rmvnorm(n_t, sigma=V_ss )

# Project through time and add mean
for( t in seq_len(n_t) ){
  if(t>1) d[t,] = rho*d[t-1,] + d[t,]
}
#d = d + 0.5

# Shape into longform data-frame and add error
Data = data.frame( expand.grid(time=1:n_t, x=1:n_x, y=1:n_y), "var"="logn", z=exp(as.vector(d)))
Data$n = tweedie::rtweedie( n=nrow(Data), mu=Data$z, phi=0.5, power=1.5 )
mean(Data$n==0)

# make mesh
mesh = fm_mesh_2d( Data[,c('x','y')] )

# fit model
mytinyVAST = tinyVAST( dsem = "logn -> logn, 1, rho",
           data = Data,
           formula = n ~ 0 + factor(time),
           spatial_graph = mesh,
           family = tweedie() )
mytinyVAST
```
The estimated values for `beta_z` then correspond to the simulated value for `rho` and `spatial_sd`.

We can compare the true densities:

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
library(sf)
data_wide = reshape( Data[,c('x','y','time','z')],
                     direction = "wide", idvar = c('x','y'), timevar = "time")
sf_data = st_as_sf( data_wide, coords=c("x","y"))
sf_grid = sf::st_make_grid( sf_data )
sf_plot = st_sf(sf_grid, st_drop_geometry(sf_data) )
plot(sf_plot, max.plot=n_t )
```

with the estimated densities:
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
Data$z_hat = predict(mytinyVAST)
data_wide = reshape( Data[,c('x','y','time','z_hat')],
                     direction = "wide", idvar = c('x','y'), timevar = "time")
sf_data = st_as_sf( data_wide, coords=c("x","y"))
sf_plot = st_sf(sf_grid, st_drop_geometry(sf_data) )
plot(sf_plot, max.plot=n_t )
```

where a scatterplot shows that they are highly correlated:
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
plot( x=Data$z, y=Data$z_hat )
```

We can then calculate the area-weighted total abundance and compare it with its true value:
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Predicted sample-weighted total
(Est = sapply( seq_len(n_t),
   FUN=\(t) integrate_output(mytinyVAST, newdata=subset(Data,time==t)) ))

# True (latent) sample-weighted total
(True = tapply( Data$z, INDEX=Data$time, FUN=sum ))

#
Index = data.frame( time=seq_len(n_t), t(Est), True )
Index$low = Index[,'Est...bias.correct.'] - 1.96*Index[,'Std..Error']
Index$high = Index[,'Est...bias.correct.'] + 1.96*Index[,'Std..Error']

#
library(ggplot2)
ggplot(Index, aes(time, Estimate)) +
  geom_ribbon(aes(ymin = low,
                  ymax = high),    # shadowing cnf intervals
              fill = "lightgrey") +
  geom_line( color = "black",
            linewidth = 1) +
  geom_point( aes(time, True), color = "red" )
```

Next, we compare this against the current version of VAST
```{r, eval=updated_vast, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
library(VAST)
settings = make_settings( purpose="index3",
                          n_x = n_x*n_y,
                          Region = "Other",
                          bias.correct = FALSE,
                          use_anisotropy = FALSE )
settings$FieldConfig['Epsilon','Component_1'] = 0
settings$FieldConfig['Omega',] = 0
settings$RhoConfig['Epsilon2'] = 4
settings$RhoConfig['Beta1'] = 3
settings$ObsModel = c(10,2)

myVAST = fit_model( settings=settings,
                 Lat_i = Data[,'y'],
                 Lon_i = Data[,'x'],
                 t_i = Data[,'time'],
                 b_i = Data[,'n'],
                 a_i = rep(1,nrow(Data)),
                 observations_LL = cbind(Lat=Data[,'y'],Lon=Data[,'x']),
                 grid_dim_km = c(100,100),
                 newtonsteps = 0,
                 loopnum = 1,
                 control = list(eval.max=100, iter.max=100, trace=0) )
myVAST
```
```{r, eval=!updated_vast, echo=FALSE, message=FALSE, include=FALSE}
myVAST = list("total_time"=NA)
```
Or with sdmTMB

```{r, eval=has_sdmTMB, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
library(sdmTMB)
mesh = make_mesh(Data, c("x","y"), n_knots=n_x*n_y )

start_time = Sys.time()
mysdmTMB = sdmTMB(
  formula = n ~ 0 + factor(time),
  data = Data,
  mesh = mesh,
  spatial = "off",
  spatiotemporal = "ar1",
  time = "time",
  family = tweedie()
)
sdmTMBtime = Sys.time() - start_time
```
```{r, eval=!has_sdmTMB, echo=FALSE, message=FALSE, include=FALSE}
sdmTMBtime = NA
```

The models all have similar runtimes
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
Times = c( "tinyVAST" = mytinyVAST$run_time,
           "VAST" = myVAST$total_time,
           "sdmTMB" = sdmTMBtime )
knitr::kable( cbind("run times (sec.)"=Times), digits=1)
```

# Delta models

We can also fit these data using a delta model

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# fit model
mydelta = tinyVAST( data = Data,
               formula = n ~ 1,
               delta_options = list(delta_formula = ~ 0 + factor(time)),
               delta_dsem = "logn -> logn, 1, rho",
               family = delta_lognormal(),
               spatial_graph = mesh )

mydelta
```

# Bivariate spatio-temporal autoregressive model

We next highlight how to specify a bivariate spatio-temporal model with a cross-laggged (vector autoregressive) interaction.

```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Simulate settings
theta_xy = 0.2
n_x = n_y = 10
n_t = 20
B = rbind( c( 0.5, -0.25),
           c(-0.1,  0.50) )

# Simulate GMRFs
R = exp(-theta_xy * abs(outer(1:n_x, 1:n_y, FUN="-")) )
d1 = mvtnorm::rmvnorm(n_t, sigma=0.2*kronecker(R,R) )
d2 = mvtnorm::rmvnorm(n_t, sigma=0.2*kronecker(R,R) )
d = abind::abind( d1, d2, along=3 )

# Project through time and add mean
for( t in seq_len(n_t) ){
  if(t>1) d[t,,] = t(B%*%t(d[t-1,,])) + d[t,,]
}

# Shape into longform data-frame and add error
Data = data.frame( expand.grid(time=1:n_t, x=1:n_x, y=1:n_y, "var"=c("d1","d2")), z=exp(as.vector(d)))
Data$n = tweedie::rtweedie( n=nrow(Data), mu=Data$z, phi=0.5, power=1.5 )

# make mesh
mesh = fm_mesh_2d( Data[,c('x','y')] )

# Define DSEM
dsem = "
  d1 -> d1, 1, b11
  d2 -> d2, 1, b22
  d2 -> d1, 1, b21
  d1 -> d2, 1, b12
  d1 <-> d1, 0, var1
  d2 <-> d2, 0, var1
"

# fit model
out = tinyVAST( dsem = dsem,
           data = Data,
           formula = n ~ 0 + var,
           spatial_graph = mesh,
           family = tweedie() )
out
```

The values for `beta_z` again correspond to the specified value for interaction-matrix `B`

We can again calculate the area-weighted total abundance and compare it with its true value:
```{r, eval=TRUE, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Predicted sample-weighted total
Est1 = sapply( seq_len(n_t), FUN=\(t) integrate_output(out, newdata=subset(Data,time==t & var=="d1")) )
Est2 = sapply( seq_len(n_t), FUN=\(t) integrate_output(out, newdata=subset(Data,time==t & var=="d2")) )

# True (latent) sample-weighted total
True = tapply( Data$z, INDEX=list("time"=Data$time,"var"=Data$var), FUN=sum )

#
Index = data.frame( expand.grid(dimnames(True)), "True"=as.vector(True) )
Index = data.frame( Index, rbind(t(Est1), t(Est2)) )
Index$low = Index[,'Est...bias.correct.'] - 1.96*Index[,'Std..Error']
Index$high = Index[,'Est...bias.correct.'] + 1.96*Index[,'Std..Error']

#
library(ggplot2)
ggplot(Index, aes( time, Estimate )) +
  facet_grid( rows=vars(var), scales="free" ) +
  geom_segment(aes(y = low,
                  yend = high,
                  x = time,
                  xend = time) ) +
  geom_point( aes(x=time, y=Estimate), color = "black") +
  geom_point( aes(x=time, y=True), color = "red" )
```

